Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web

### 2025-05-09 autovi_paper.qmd
Add an extension to GLM Poisson

The `autovi` package is available on CRAN. It is actively developed and maintained, with the latest updates accessible on GitHub. This paper uses `autovi` version 0.4.1.
The package includes internal functions to check the current Python environment used by the `reticulate` package. If the necessary Python packages are not installed in the Python interpreter, an error will be raised. If you want to select a specific Python environment, you can do so by calling the `reticulate::use_python()` function before using the `autovi` package.

@@

The modules for VSS prediction and $p$-value computation are predefined and cannot be overridden, although users can interact with them directly through function arguments. Similarly, the image reading and resizing module is fixed but will adapt to different Keras models by checking their input shapes. The remaining seven modules are designed to be overridable, enabling users to tailor the infrastructure to their specific needs. These modules are discussed in detail in the package documentation. <!-- in the following sections.-->
<!-- FIXME: ET: which website? >
<!--
### Initialization

>>>>

The `autovi` package is available on CRAN. It is actively developed and maintained, with the latest updates accessible on GitHub. This paper uses `autovi` version 0.4.2.
The package includes internal functions to check the current Python environment used by the `reticulate` package. If the necessary Python packages are not installed in the Python interpreter, an error will be raised. If you want to select a specific Python environment, you can do so by calling the `reticulate::use_python()` function before using the `autovi` package.

@@

The modules for VSS prediction and $p$-value computation are predefined and cannot be overridden, although users can interact with them directly through function arguments. Similarly, the image reading and resizing module is fixed but will adapt to different Keras models by checking their input shapes. The remaining seven modules are designed to be overridable, enabling users to tailor the infrastructure to their specific needs. These modules are discussed in detail in the package documentation.

## Extension to Other Model Classes

The `autovi` R package can be extended to accommodate other classes of models beyond linear regression, such as generalized linear models (`glm`). This is achieved by substituting the relevant overridable modules, as illustrated in @fig-autovi-diag.

We provide an example of defining a new checker class tailored for Poisson regression using the `glm` framework:

1. Define a new class using `new_class()` with `AUTO_VI` as the parent class.
2. Override the necessary methods using `register_method()`. In this example, we use Pearson residuals. To simulate null residuals, we assume the fitted model is correct and the estimated coefficients are accurate. New response values are generated accordingly, and a new model is fitted to this simulated response. Null residuals are then extracted from this refitted model.
3. Create an alias for the `instantiate()` method of the new class.

```{r echo = TRUE}
AUTO_GLM_VI <- new_class(AUTO_VI, class_name = "AUTO_GLM_VI")
register_method(
  AUTO_GLM_VI,
  get_fitted_and_resid = function(fitted_model = self$fitted_model) {
    tibble(.fitted = fitted(fitted_model),
           .resid = resid(fitted_model, type = "pearson"))
  },
  null_method = function(fitted_model = self$fitted_model) {
    dat <- model.frame(fitted_model)
    dat[[1]] <- rpois(nrow(dat), lambda = fitted(fitted_model))
    new_mod <- update(fitted_model, data = dat)
    return(self$get_fitted_and_resid(new_mod))
  }
)
auto_glm_vi <- AUTO_GLM_VI$instantiate
```

The resulting checker class can be employed analogously to the linear model case described in @sec-autovi-usage. For illustration, we fit a Poisson model in which the quadratic term of the predictor $x$ is intentionally omitted. This misspecification manifests as a pronounced U-shaped pattern in the lineup display, which is also successfully identified by the computer vision model, yielding a p-value substantially below the conventional threshold of 0.05.

It is important to note, however, that the pre-trained computer vision model was developed specifically for diagnostic evaluation of linear regression. Its applicability to other model classes relies on the assumption that the null residual plots exhibit characteristics broadly consistent with those of well-behaved linear regression residuals, that is, residuals should be approximately randomly scattered around zero, display roughly constant variance across the range of fitted values, and exhibit no discernible structure or curvature. If these conditions are not met, or if model violations do not give rise to visually detectable patterns, the validity of the automated diagnostics may be compromised. For further discussion on extending the methodology to other model classes, refer to @li2024automated.

```r
set.seed(10086)
x <- rnorm(300, sd = 0.5)
y <- rpois(300, lambda = exp(1 + x + x^2))
y_fit <- glm(y ~ x, family = "poisson")

glm_checker <- auto_glm_vi(glm(y ~ x, family = "poisson"),
                           keras_model = get_keras_model("vss_phn_32"))
glm_checker$plot_lineup()
```

```{r include = FALSE}
set.seed(10086)
x <- rnorm(300, sd = 0.5)
y <- rpois(300, lambda = exp(1 + x + x^2))
y_fit <- glm(y ~ x, family = "poisson")

glm_checker <- auto_glm_vi(glm(y ~ x, family = "poisson"))
```

```{r}
glm_checker$plot_lineup()
```

```r
glm_checker$check()
```

```{r message = TRUE}
if (!file.exists("cached_data/autovi_glm_checker_demo_print.rds")) {
  glm_checker <- auto_glm_vi(glm(y ~ x, family = "poisson"),
                           keras_model = get_keras_model("vss_phn_32"))
  glm_checker$check()
  saveRDS(glm_checker$..str..(), "cached_data/autovi_glm_checker_demo_print.rds")
} else {
  cli::cli_h3(readRDS("cached_data/autovi_glm_checker_demo_print.rds"))
}
```

<!-- in the following sections.-->
<!-- FIXME: ET: which website? >
<!--
### Initialization

### 2025-05-09 autovi_paper.tex
Add an extension to GLM Poisson

The \texttt{autovi} package is available on CRAN. It is actively
developed and maintained, with the latest updates accessible on GitHub.
This paper uses \texttt{autovi} version 0.4.1. The package includes
internal functions to check the current Python environment used by the
\texttt{reticulate} package. If the necessary Python packages are not
installed in the Python interpreter, an error will be raised. If you
want to select a specific Python environment, you can do so by calling
the \texttt{reticulate::use\_python()} function before using the
\texttt{autovi} package.

@@

>>>>

The \texttt{autovi} package is available on CRAN. It is actively
developed and maintained, with the latest updates accessible on GitHub.
This paper uses \texttt{autovi} version 0.4.2. The package includes
internal functions to check the current Python environment used by the
\texttt{reticulate} package. If the necessary Python packages are not
installed in the Python interpreter, an error will be raised. If you
want to select a specific Python environment, you can do so by calling
the \texttt{reticulate::use\_python()} function before using the
\texttt{autovi} package.

@@

\subsection{Extension to Other Model
Classes}\label{extension-to-other-model-classes}

The \texttt{autovi} R package can be extended to accommodate other
classes of models beyond linear regression, such as generalized linear
models (\texttt{glm}). This is achieved by substituting the relevant
overridable modules, as illustrated in Figure~\ref{fig-autovi-diag}.

We provide an example of defining a new checker class tailored for
Poisson regression using the \texttt{glm} framework:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define a new class using \texttt{new\_class()} with \texttt{AUTO\_VI}
  as the parent class.
\item
  Override the necessary methods using \texttt{register\_method()}. In
  this example, we use Pearson residuals. To simulate null residuals, we
  assume the fitted model is correct and the estimated coefficients are
  accurate. New response values are generated accordingly, and a new
  model is fitted to this simulated response. Null residuals are then
  extracted from this refitted model.
\item
  Create an alias for the \texttt{instantiate()} method of the new
  class.
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{AUTO\_GLM\_VI }\OtherTok{\textless{}{-}} \FunctionTok{new\_class}\NormalTok{(AUTO\_VI, }\AttributeTok{class\_name =} \StringTok{"AUTO\_GLM\_VI"}\NormalTok{)}
\FunctionTok{register\_method}\NormalTok{(}
\NormalTok{  AUTO\_GLM\_VI,}
  \AttributeTok{get\_fitted\_and\_resid =} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{fitted\_model =}\NormalTok{ self}\SpecialCharTok{$}\NormalTok{fitted\_model) \{}
    \FunctionTok{tibble}\NormalTok{(}\AttributeTok{.fitted =} \FunctionTok{fitted}\NormalTok{(fitted\_model),}
           \AttributeTok{.resid =} \FunctionTok{resid}\NormalTok{(fitted\_model, }\AttributeTok{type =} \StringTok{"pearson"}\NormalTok{))}
\NormalTok{  \},}
  \AttributeTok{null\_method =} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{fitted\_model =}\NormalTok{ self}\SpecialCharTok{$}\NormalTok{fitted\_model) \{}
\NormalTok{    dat }\OtherTok{\textless{}{-}} \FunctionTok{model.frame}\NormalTok{(fitted\_model)}
\NormalTok{    dat[[}\DecValTok{1}\NormalTok{]] }\OtherTok{\textless{}{-}} \FunctionTok{rpois}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(dat), }\AttributeTok{lambda =} \FunctionTok{fitted}\NormalTok{(fitted\_model))}
\NormalTok{    new\_mod }\OtherTok{\textless{}{-}} \FunctionTok{update}\NormalTok{(fitted\_model, }\AttributeTok{data =}\NormalTok{ dat)}
    \FunctionTok{return}\NormalTok{(self}\SpecialCharTok{$}\FunctionTok{get\_fitted\_and\_resid}\NormalTok{(new\_mod))}
\NormalTok{  \}}
\NormalTok{)}
\NormalTok{auto\_glm\_vi }\OtherTok{\textless{}{-}}\NormalTok{ AUTO\_GLM\_VI}\SpecialCharTok{$}\NormalTok{instantiate}
\end{Highlighting}
\end{Shaded}

The resulting checker class can be employed analogously to the linear
model case described in Section~\ref{sec-autovi-usage}. For
illustration, we fit a Poisson model in which the quadratic term of the
predictor \(x\) is intentionally omitted. This misspecification
manifests as a pronounced U-shaped pattern in the lineup display, which
is also successfully identified by the computer vision model, yielding a
p-value substantially below the conventional threshold of 0.05.

It is important to note, however, that the pre-trained computer vision
model was developed specifically for diagnostic evaluation of linear
regression. Its applicability to other model classes relies on the
assumption that the null residual plots exhibit characteristics broadly
consistent with those of well-behaved linear regression residuals, that
is, residuals should be approximately randomly scattered around zero,
display roughly constant variance across the range of fitted values, and
exhibit no discernible structure or curvature. If these conditions are
not met, or if model violations do not give rise to visually detectable
patterns, the validity of the automated diagnostics may be compromised.
For further discussion on extending the methodology to other model
classes, refer to \citet{li2024automated}.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{10086}\NormalTok{)}
\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{300}\NormalTok{, }\AttributeTok{sd =} \FloatTok{0.5}\NormalTok{)}
\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{rpois}\NormalTok{(}\DecValTok{300}\NormalTok{, }\AttributeTok{lambda =} \FunctionTok{exp}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x }\SpecialCharTok{+}\NormalTok{ x}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{))}
\NormalTok{y\_fit }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{family =} \StringTok{"poisson"}\NormalTok{)}

\NormalTok{glm\_checker }\OtherTok{\textless{}{-}} \FunctionTok{auto\_glm\_vi}\NormalTok{(}\FunctionTok{glm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{family =} \StringTok{"poisson"}\NormalTok{),}
                           \AttributeTok{keras\_model =} \FunctionTok{get\_keras\_model}\NormalTok{(}\StringTok{"vss\_phn\_32"}\NormalTok{))}
\NormalTok{glm\_checker}\SpecialCharTok{$}\FunctionTok{plot\_lineup}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{center}
\includegraphics[width=1\textwidth,height=\textheight]{autovi_paper_files/figure-pdf/unnamed-chunk-9-1.pdf}
\end{center}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{glm\_checker}\SpecialCharTok{$}\FunctionTok{check}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
-- <AUTO_GLM_VI object>
Status:
 - Fitted model: glm, lm
 - Keras model: (None, 32, 32, 3) + (None, 5) -> (None, 1)
    - Output node index: 1
 - Result:
    - Observed visual signal strength: 4.875 (p-value = 0.009901)
    - Null visual signal strength: [100 draws]
       - Mean: 1.331
       - Quantiles: 
          ╔══════════════════════════════════════════╗
          ║  25%   50%   75%   80%   90%   95%   99% ║
          ║1.035 1.233 1.488 1.644 1.941 2.276 2.639 ║
          ╚══════════════════════════════════════════╝
    - Bootstrapped visual signal strength: [100 draws]
       - Mean: 5.51 (p-value = 0.009901)
       - Quantiles: 
          ╔══════════════════════════════════════════╗
          ║  25%   50%   75%   80%   90%   95%   99% ║
          ║5.330 5.505 5.698 5.735 5.830 5.903 6.013 ║
          ╚══════════════════════════════════════════╝
    - Likelihood ratio: 0.05096 (boot) / 0 (null) = Extremely large 
\end{verbatim}
